Analysing the Factors Influencing the Wages of the US Workforce in 2023

### 2024-12-06 paper/paper.qmd
Final commit

Income inequality is a persistent social and economic problem with significant implications for individual well-being and societal stability. Understanding how personal characteristics and external factors shape income distribution is essential to addressing its underlying causes. While many studies have explored income inequality, few have systematically analyzed the interaction of these factors. Additionally, existing research often lacks recent data that are both nationally representative and adaptable for diverse analyses. This study addresses these gaps using the 2023 sample from the IPUMS database, which allows users to customize variables and provides high-quality, comprehensive data. Leveraging this resource, the study identifies the main determinants of income inequality through a multiple linear regression model.

@@

This paper provides a comprehensive analysis of income inequality by examining its key drivers and impacts. @sec-data describes the dataset in detail, highlights its main features, and uses visualizations to present the analysis results. @sec-model focuses on the construction of the multiple linear regression model, validates its effectiveness, and presents the results of the analysis. @sec-discussion analyzes the research findings, explores the reasons behind the observed patterns, and offers policy recommendations to address income inequality. It also identifies the study’s limitations and suggests directions for future research. @sec-appendix provides additional details on the model, and methods of survey and sampling, offering more comprehensive support for the study.

@@

This study uses sample data from the 2023 American Community Survey (ACS). The dataset is based on a 1% random national sample, ensuring broad representativeness. The sample includes individuals living in private households and those in group quarters, such as student dormitories, correctional facilities, and care homes. Weights are applied to the data to reduce biases caused by the sampling design and nonresponse. This improves the generalizability of the results. The smallest geographic unit in the dataset is the Public Use Microdata Area (PUMA). Each PUMA includes at least 100,000 residents and is contained within state boundaries. This design anonymizes respondents’ locations while enabling regional analysis.

The ACS collects data using multiple methods. These include mailed questionnaires, online submissions, and in-person or telephone interviews. The surveys cover a wide range of topics, such as demographics, education, employment, and income. For example, wage income is recorded by asking respondents to report their total pre-tax earnings from wages, salaries, tips, and other sources over the past 12 months. Similarly, the variable UHRSWORK (usual hours worked per week) measures labor input based on respondents’ self-reported average weekly work hours during the survey.

Despite its robust sampling methodology and extensive coverage, the ACS data has some limitations. Variables such as income and work hours rely heavily on self-reported responses. This introduces potential inaccuracies due to recall bias, rounding errors, or deliberate misreporting. For individuals with irregular or multiple income sources, questions about wages or hours worked may be interpreted differently, further increasing measurement errors.

To address these issues, the ACS uses data validation checks and imputation techniques to handle incomplete or inconsistent responses. However, these methods cannot fully eliminate the risk of bias or error. Even with these limitations, the ACS dataset remains a reliable and widely used resource for studying wage income and related factors. Its combination of sampling weights, large sample size, and rigorous design framework supports robust analysis.

@@

@fig-demographics-distribution illustrates the distribution of demographic characteristics and their relationship with average income across gender, race, region, and education level. Pie charts display the proportional distribution of each group, while bar charts show corresponding average income levels, providing a straightforward depiction of these relationships.

Chart (a) shows the gender distribution: females account for 48.9% of the population, and males 51.1%, indicating a nearly equal split. Chart (b) shows that males earn an average income of $82,840, compared to $57,146 for females.

Chart (c) presents the racial composition, with White individuals forming the largest group at 67%, followed by Black individuals at 18.2%, Asians at 6.8%, and other racial groups at 8.1%. Chart (d) shows that Asians have the highest average income at $96,574, followed by White individuals at $73,778. Black individuals and other racial groups earn $51,221 and $56,604, respectively.

Chart (e) shows the regional distribution. The West represents the largest share at 29%, followed by the Northeast (26.6%), the South (24.1%), and the Midwest (20.3%). Chart (f) indicates that the Midwest has the lowest average income at $62,046, while the Northeast has the highest at $78,138. The South and West have average incomes of $66,954 and $75,293, respectively.

Chart (g) illustrates the distribution of education levels: 16.5% of the population has less than a high school education, 30.4% has completed high school, 23.5% has some college education, 26.2% holds a bachelor’s degree, and 9.5% has education beyond a bachelor’s degree. Chart (h) highlights the strong relationship between education and income. Those with less than a high school education earn an average of $36,149, while individuals with a bachelor’s degree earn $87,465, which is approximately 2.4 times higher. Income continues to increase for those with education beyond a bachelor’s degree, averaging $124,144, further underscoring the significant impact of education on earnings.

@@

@fig-combined-visualizations examines the relationship between income, work hours, and age, incorporating both individual-level data and aggregated trends.

Chart (a) visualizes the relationship between work hours and income, where the red quadratic regression line indicates a positive correlation. Longer work hours generally correspond to higher income, with the regression line reflecting a stable linear trend rather than noticeable deceleration. Income variability increases with longer work hours, particularly beyond 50 hours per week, where fluctuations become more pronounced. Chart (b) represents the trend of average income by work hours. Income shows a steady increase with work hours, especially between 20 and 50 hours per week, but becomes less consistent beyond 50 hours.

Chart (c) illustrates the relationship between age and income, with the red quadratic regression line highlighting a pattern where income rises with age before gradually declining. Chart (d) presents average income by age as a line chart. It shows that average income steadily increases with age, peaking in the late 40s and early 50s, followed by a gradual decline as individuals approach retirement age.

@@

@fig-Education-Race illustrates the proportion of different education levels across racial groups. The intensity of blue indicates the proportion, with darker shades representing higher values.

Among Asians, the proportion of highly educated individuals is relatively high, with 38% holding a bachelor's degree and 30% having education beyond a bachelor's degree, while the proportion of those with "Below High School" education is 0%. In the Black group, "High School" accounts for the largest proportion at 32%, followed by "Some College" at 27%, while those with "Above Bachelor" degrees make up only 16%. In the White group, "High School" accounts for 29%, with 27% holding a bachelor's degree and 24% having "Some College" education, while 17% have "Above Bachelor" degrees. Among the "Other" group, 40% have a "High School" education, 25% have "Some College," 17% hold a bachelor's degree, and only 9% have "Above Bachelor" degrees.

Educational attainment shows significant variation across racial groups, reflecting the complex relationship between race and educational background. Asians have the highest levels of education, with a strong representation in bachelor's and advanced degrees. In contrast, Black and White groups display a broader distribution, with high school education being more common. The "Other" category, encompassing smaller racial groups and mixed-race individuals, stands out with a concentration at the high school level and lower representation in higher education, suggesting potential socio-economic and cultural disparities.

@@

The goal of our modelling strategy is to quantify the contributions of key factors—such as weekly working hours, age (and its quadratic term), education level, region, gender, and race—to variations in income (log-transformed). By estimating the relative impact of these predictors, the model provides insights into potential social and demographic disparities in income.

@@

Multiple regression models rely on several assumptions: linearity, meaning a linear relationship exists between independent and dependent variables; homoskedasticity, requiring constant error variance across predictors; independence of errors, meaning residuals are uncorrelated; normality, where residuals follow a normal distribution; and independence of independent variables, ensuring no significant multicollinearity. These assumptions are discussed in @sec-model-details and evaluated using the diagnostic plots shown in @fig-check.

@@

The selection of variables and model structure was based on theory and data characteristics. Given the right-skewed nature of the income distribution in @fig-income-distribution, the dependent variable is log-transformed (`log(INCWAGE)`) to improve model fit. Furthermore, to capture the nonlinear relationship between age and income evident in @fig-combined-visualizations, a quadratic term for age (`age^2`) is included in the model. Categorical variables, such as region, education level, race group, and gender, are represented using dummy variables. Each category is compared to a reference category, allowing the regression model to capture and interpret group-level differences.

During the model selection process, an initial model included an interaction term between education level and race group (education_level:race_group), based on the association observed in @fig-Education-Race, while other variables remained unchanged. This interaction term aimed to evaluate income differences across racial groups at the same education level. However, most interaction terms were not statistically significant and provided minimal improvement to model fit metrics, such as adjusted $R^2$ and AIC. To simplify the model and enhance interpretability, this interaction term was excluded from the final specification.

The evaluation of the final model's performance was conducted on both training and testing datasets. As shown in @tbl-test, the RMSE and MAE values are slightly higher for the testing data compared to the training data (0.786 vs. 0.773 for RMSE, and 0.568 vs. 0.543 for MAE), reflecting a modest decline in predictive accuracy when applied to unseen data. Similarly, the $R^2$ values are 0.488 for the training set and 0.468 for the testing set, indicating that the model captures a moderate proportion of the variability in the log-transformed income. Additionally, all variables included in the final model were statistically significant, with $p$-values below 0.05, further supporting the robustness of the selected predictors. Overall, the final model achieves a practical balance between complexity, interpretability, and predictive performance.

@@

@fig-coefficients shows the effects of weekly working hours, age, region, education level, gender, and racial group on income. The coefficient plot also displays the 95% confidence intervals for each coefficient, with detailed numerical values presented in @tbl-modelresults in @sec-model-details.

Education level is the most important factor. Individuals with lower education, such as "Below High School," have negative coefficients, while those with higher education, such as a "Bachelor’s" degree, show positive coefficients. This highlights the significant economic benefits of higher education. However, the confidence intervals for education vary. For example, "Below High School" has a wider interval, indicating greater uncertainty in the estimate.

Age also has a strong impact on income. The positive coefficient for age shows that income increases as people get older. However, the negative coefficient for age squared suggests that this growth slows over time and may decline later in life. The confidence interval for age is narrow, indicating a precise estimate. Gender differences are also evident. Male individuals have higher income, as shown by the positive coefficient for gender. The narrow confidence interval for gender confirms the reliability of this effect.

Racial factor present more complex dynamics. For example, "Black" and "Other" groups have negative coefficients and wider confidence intervals, reflecting greater uncertainty in their effects. In contrast, "White" individuals have coefficients close to zero and narrow intervals, showing minimal differences from the reference group. Regional effects are also notable. The "West" has positive coefficients compared to the reference region, while other regions show smaller or slightly negative effects. Some regional variables have confidence intervals that include zero, suggesting these effects may not be statistically significant.

Finally, the effect of weekly working hours on income is relatively small but still positive. Its narrow confidence interval indicates a stable and reliable estimate. Overall, these results provide a clear understanding of the socioeconomic factors influencing income and offer valuable insights for future research and policy discussions.

@@

This paper uses a multiple linear regression model to analyze key factors that influence income, including education, age, gender, race, region, and hours worked. The study is based on data from the 2023 American Community Survey (ACS), cleaned and processed for visualization and analysis. The model quantifies the impact of these variables and explores their significance, showing how personal characteristics and external factors shape the income distribution. Education, age, gender, and race are treated as endogenous variables, reflecting individual attributes, while region and work hours are considered exogenous, capturing the influence of external conditions on income. Through the analysis of these variables, the study provides insights into the roots of income inequality and offers evidence to inform policymaking.

## Education and Age: Core Drivers of Income

Education is a key factor influencing income, with a clear positive correlation between education level and earnings. Graduate degree holders earn significantly more, while a bachelor's degree also boosts income. However, those with only 1-2 years of college earn similar incomes to high school graduates, underscoring the importance of completing college. Higher education plays a crucial role in enhancing economic potential and securing high-paying jobs. Most income differences stem from variations in bachelor's, graduate, and doctoral education levels, while pre-college education has a smaller impact.

Age also has a nonlinear impact on income. The model indicates that income peaks between ages 40 and 50 before gradually declining. The quadratic term captures this trend, reflecting how different stages of a career affect earnings. Early in their careers, individuals experience rapid income growth as they gain experience and promotions. In later stages, productivity declines, and income typically decreases as retirement approaches.

Education and age shape an individual's earning potential. These findings are valuable for career planning and offer guidance for policymakers. Individuals can enhance their competitiveness in the job market by pursuing a bachelor's degree or higher. Governments can increase education funding and provide financial aid to expand access to higher education. Additionally, support programs for those nearing retirement can help mitigate the impact of declining wages on their quality of life.

@@

Gender inequality in income distribution is evident. Studies show that men earn significantly more than women, even when controlling for education, age, and other factors. This disparity may stem from gender bias and occupational segregation. Women are often expected to take on more family responsibilities and face potential career interruptions due to childbirth [@gender]. Additionally, men are overrepresented in many high-paying professions, contributing to lower average wages for women [@blau2013trends].

Racial disparities, on the other hand, highlight another form of structural inequality. Research indicates that Asians earn the highest incomes, followed by Whites, while Black individuals and those from "other" racial groups earn the least. These differences can be attributed to historical and systemic factors, such as racial discrimination in the labor market and the underrepresentation of minorities in high-paying industries [@race]. Educational disparities between racial groups, as shown in @fig-Education-Race, also contribute to these income gaps.

Gender and race are exogenous factors, beyond individual control, and the wage disparities they cause reflect systemic inequality. Addressing the gender pay gap requires targeted policies, such as promoting pay transparency and increasing women's participation in high-paying industries. To reduce racial income disparities, policymakers can implement measures like providing education and employment support for minority groups, encouraging diversity hiring, and strictly enforcing anti-discrimination laws. Cultural change is also crucial; reducing biases against women and racial groups can help narrow wage gaps and promote fairness.

@@

Work hours are positively linked to income, but the returns diminish. Beyond 50 hours per week, income becomes more unstable. This suggests that working longer hours does not always lead to higher pay. Policies like overtime pay rules or limits on work hours may play a role. The government should introduce better labor laws, such as capping excessive work hours. Encouraging businesses to innovate and improve efficiency can reduce reliance on long hours. These measures can protect workers’ rights while promoting a fairer and healthier labor market.

@@

The data used in this study has certain limitations. It relies on survey responses, which can include self-reporting bias. For example, respondents may overestimate or underestimate their income or working hours. Additionally, the original data collection only classified gender as male or female. Non-binary and transgender groups were not included, leaving some populations unrepresented. Future research could incorporate more diverse datasets, such as administrative records or third-party data, to improve accuracy. Using more detailed classifications could also better reflect overlooked groups and enhance inclusivity.

Some important variables were not included in the analysis, such as industry classification. While industry significantly affects income, its complex structure and lack of standardized categories made it difficult to analyze. This exclusion might have introduced bias into the results. To simplify the model, smaller racial groups, such as Indigenous, and multiracial individuals, were categorized as “Other.” Although this made the analysis manageable, it overlooked finer details in racial differences. Future studies could refine these categories and include more detailed variables when sample sizes allow.

>>>>

Income inequality is a persistent social and economic problem with significant implications for individual well-being and societal stability. Understanding how personal characteristics and external factors shape income distribution is essential to addressing its underlying causes. While many studies have explored income inequality, few have systematically analyzed the interaction of these factors. Additionally, existing research often lacks recent data that are both nationally representative and adaptable for diverse analyses. This study addresses these gaps using the 2023 sample from the IPUMS database, which allows users to customize variables and provides high-quality, detailed data. Using this resource, the study identifies the main determinants of income inequality through a multiple linear regression model.

@@

This paper provides a thorough analysis of income inequality by examining its main factors and effects. @sec-data describes the dataset in detail, highlights its main features, and uses visualizations to present the analysis results. @sec-model focuses on the construction of the multiple linear regression model, validates its effectiveness, and presents the results of the analysis. @sec-discussion analyzes the research findings, explores the reasons behind the observed patterns, and offers policy recommendations to address income inequality. It also identifies the study’s limitations and suggests directions for future research. @sec-appendix provides additional details on the model, and methods of survey and sampling, offering more detailed support for the study.

@@

This study uses sample data from the 2023 American Community Survey (ACS). The dataset is based on a 1% random national sample, ensuring broad representativeness. The sample includes individuals living in private households and those in group quarters, such as student dormitories, correctional facilities, and care homes. Weights are applied to reduce biases from the sampling design and nonresponse, enhancing result generalizability. The smallest geographic unit in the dataset is the Public Use Microdata Area (PUMA). Each PUMA includes at least 100,000 residents and is contained within state boundaries. This design anonymizes respondents’ locations while enabling regional analysis.

The ACS collects data through mailed questionnaires, online submissions, and in-person or telephone interviews. The surveys cover a wide range of topics, such as demographics, education, employment, and income. For example, wage income is recorded by asking respondents to report their total pre-tax earnings from wages, salaries, tips, and other sources over the past 12 months. Similarly, the variable UHRSWORK (usual hours worked per week) measures labor input based on respondents’ self-reported average weekly work hours during the survey.

Despite its robust sampling methodology and broad coverage, the ACS data has limitations. Variables such as income and work hours rely heavily on self-reported responses, introducing potential inaccuracies due to recall bias, rounding errors, or deliberate misreporting. For individuals with irregular or multiple income sources, questions about wages or hours worked may be interpreted variably, increasing measurement errors.

To address these issues, the ACS uses data validation checks and imputation techniques to handle incomplete or inconsistent responses. However, these methods cannot fully eliminate the risk of bias or error. Despite these limitations, the ACS dataset remains a reliable resource for analyzing wage income and related factors. Its combination of sampling weights, large sample size, and rigorous design framework supports robust analysis.

@@

@fig-demographics-distribution illustrates the distribution of demographic characteristics and their relationship with average income across gender, race, region, and education level. Group proportions are shown alongside their corresponding average income levels, clearly illustrating these relationships.

Chart (a) illustrates gender distribution: females make up 48.9% and males 51.1% of the population, indicating a near-equal split. Chart (b) shows males with an average income of $82,840, compared to $57,146 for females.

Chart (c) presents the racial composition, with White individuals forming the largest group at 67%, followed by Black individuals at 18.2%, Asians at 6.8%, and other racial groups at 8.1%. Chart (d) shows Asians with the highest average income at $96,574, followed by White individuals at $73,778. Black individuals and other racial groups earn $51,221 and $56,604, respectively.

Chart (e) illustrates regional distribution. The West represents the largest share at 29%, followed by the Northeast (26.6%), the South (24.1%), and the Midwest (20.3%). Chart (f) indicates that the Midwest has the lowest average income at $62,046, while the Northeast has the highest at $78,138. The South and West have average incomes of $66,954 and $75,293, respectively.

Chart (g) illustrates education levels: 16.5% have less than a high school education, 30.4% completed high school, 23.5% attended some college, 26.2% hold a bachelor’s degree, and 9.5% pursued education beyond a bachelor’s degree. Chart (h) highlights the strong relationship between education and income. Those with less than a high school education earn an average of $36,149, while individuals with a bachelor’s degree earn $87,465, approximately 2.4 times higher. Income rises further for those with education beyond a bachelor’s degree, averaging $124,144, highlighting the substantial impact of education on earnings.

@@

```
@fig-combined-visualizations examines the relationship between income, work hours, and age, incorporating both individual-level data and aggregated trends.

Chart (a) shows the relationship between work hours and income, with the red quadratic regression line indicating a positive correlation. Longer work hours generally correspond to higher income, with the regression line reflecting a stable linear trend without noticeable deceleration. Income variability grows with longer work hours, especially beyond 50 hours per week, where fluctuations are more pronounced. Chart (b) depicts average income trends across work hours. Income steadily increases with work hours, especially between 20 and 50 hours per week, but becomes less consistent beyond 50 hours.

Chart (c) shows the relationship between age and income, with the red quadratic regression line revealing a rise in income with age before a gradual decline. Chart (d) shows average income by age as a line graph. It reveals that average income steadily increases with age, peaking in the late 40s and early 50s, followed by a gradual decline as individuals approach retirement age.

@@

@fig-Education-Race illustrates the proportion of different education levels across racial groups. Darker shades of blue indicate higher proportions.

Among Asians, 38% hold a bachelor's degree and 30% have education beyond it, while none fall under "Below High School." For Black individuals, "High School" is most common at 32%, followed by "Some College" at 27%, and "Above Bachelor" at 16%. In the White group, 29% have a "High School" education, 27% hold a bachelor's degree, 24% have "Some College" education, and 17% have "Above Bachelor" degrees. The "Other" category shows 40% with a "High School" education, 25% with "Some College," 17% with a bachelor's degree, and only 9% with "Above Bachelor" degrees.

Educational attainment varies significantly across racial groups, reflecting the nuanced relationship between race and education. Asians exhibit the highest levels of education, with strong representation in bachelor's and advanced degrees. In contrast, Black and White groups show broader distributions, with high school education being more common. The "Other" category, which includes smaller racial groups and mixed-race individuals, is concentrated at the high school level and underrepresented in higher education, indicating possible socio-economic and cultural disparities.

@@

The goal of this modelling strategy is to quantify the contributions of factors such as weekly working hours, age (including its quadratic term), education level, region, gender, and race to variations in log-transformed income. By estimating the relative impact of these predictors, the model examines potential social and demographic disparities in income.

@@

Multiple regression models rely on several assumptions: linearity, indicating a linear relationship between predictors and the dependent variable; homoskedasticity, requiring constant error variance across predictors; independence of errors, where residuals are uncorrelated; normality, where residuals follow a normal distribution; and low multicollinearity among predictors. These assumptions are discussed in @sec-model-details and evaluated using the diagnostic plots shown in @fig-check.

@@

The selection of variables and model structure was based on theory and data characteristics. To address the right-skewed income distribution shown in @fig-income-distribution, the dependent variable is log-transformed (log(INCWAGE)) to improve model fit. Additionally, a quadratic term for age (age^2) is included to capture the nonlinear relationship between age and income observed in @fig-combined-visualizations. Categorical variables, such as region, education level, race group, and gender, are represented using dummy variables. Each category is compared to a reference group, enabling the model to interpret group-level differences.

During the model selection process, an initial model included an interaction term between education level and race group (education_level:race_group), based on the association observed in @fig-Education-Race, while other variables remained unchanged. This interaction term aimed to evaluate income differences across racial groups at the same education level. However, most interaction terms were not statistically significant and contributed minimally to metrics like adjusted $R^2$ and AIC. To simplify the model and enhance interpretability, this interaction term was excluded from the final specification.

The final model's performance was evaluated on both training and testing datasets. As shown in @tbl-test, the RMSE and MAE values are slightly higher for the testing data compared to the training data (0.786 vs. 0.773 for RMSE, and 0.568 vs. 0.543 for MAE), reflecting a modest decline in predictive accuracy when applied to unseen data. Similarly, the $R^2$ values of 0.488 for the training set and 0.468 for the testing set indicate that the model captures a moderate share of the variability in log-transformed income. Additionally, all variables included in the final model were statistically significant, with $p$-values below 0.05, supporting the robustness of the selected predictors. Overall, the final model achieves a practical balance between complexity, interpretability, and predictive performance.

@@

@fig-coefficients shows the effects of weekly working hours, age, region, education level, gender, and racial group on income. The plot displays 95% confidence intervals for each coefficient, with detailed numerical values provided in @tbl-modelresults in @sec-model-details.

Education level is the most important factor. Individuals with lower education, such as "Below High School," have negative coefficients, while those with higher education, such as a "Bachelor’s" degree, show positive coefficients. This underscores the economic advantages of higher education. However, the confidence intervals for education vary. For example, "Below High School" has a wider interval, indicating greater uncertainty in the estimate.

Age significantly influences income. The positive coefficient for age shows that income increases as people get older. However, the negative coefficient for age squared suggests that this growth slows over time and may decline later in life. The confidence interval for age is narrow, indicating a precise estimate. Gender differences are also evident. Male individuals have higher income, as shown by the positive coefficient for gender. The narrow confidence interval for gender confirms the reliability of this effect.

Racial factors present more complex dynamics. For example, "Black" and "Other" groups have negative coefficients and wider confidence intervals, reflecting greater uncertainty in their effects. In contrast, "White" individuals have coefficients close to zero and narrow intervals, showing minimal differences from the reference group. Regional effects are also notable. The "West" has positive coefficients compared to the reference region, while other regions show smaller or slightly negative effects. Some regional variables have confidence intervals that include zero, suggesting these effects may not be statistically significant.

Finally, the effect of weekly working hours on income is relatively small but still positive. Its narrow confidence interval indicates a stable and reliable estimate. Overall, these results clarify the socioeconomic factors influencing income and offer meaningful implications for future research and policy discussions.

@@

This paper uses a multiple linear regression model to analyze factors influencing income, including education, age, gender, race, region, and hours worked. This study uses data from the 2023 American Community Survey (ACS), which was cleaned and processed for analysis. The model quantifies the impact of these variables and explores their significance, showing how personal characteristics and external factors shape income distribution. Education, age, gender, and race are treated as endogenous variables, reflecting individual attributes, while region and work hours are considered exogenous, capturing the influence of external conditions on income. By analyzing these variables, the study provides evidence to inform policymaking and examines the roots of income inequality.

## Education and Age: Primary Determinants of Income

Education strongly influences income, with a clear positive correlation between education level and earnings. Graduate degree holders earn significantly more, while a bachelor's degree also boosts income. However, those with only 1-2 years of college earn similar incomes to high school graduates, underscoring the importance of completing college. Higher education enhances economic potential and secures high-paying jobs. Most income differences stem from variations in bachelor's, graduate, and doctoral education levels, while pre-college education has a smaller impact.

Age significantly affects income. The model indicates that income peaks between ages 40 and 50 before gradually declining. The quadratic term captures this trend, reflecting how career stages impact earnings. Early in their careers, individuals see rapid income growth with experience and promotions. In later stages, productivity declines, and income typically decreases as retirement approaches.

Education and age shape an individual's earning potential. These findings inform career planning and provide guidance for policymakers. Individuals can enhance their competitiveness in the job market by pursuing a bachelor's degree or higher. Governments can increase education funding and provide financial aid to expand access to higher education. Additionally, support programs for those nearing retirement can help mitigate the impact of declining wages on their quality of life.

@@

Income distribution shows clear gender inequality. Studies show that men earn significantly more than women, even when controlling for education, age, and other factors. This disparity may stem from gender bias and occupational segregation. Women are often expected to take on more family responsibilities and face potential career interruptions due to childbirth [@gender]. Additionally, men are overrepresented in many high-paying professions, contributing to lower average wages for women [@blau2013trends].

Racial disparities underscore another form of structural inequality. Research indicates that Asians earn the highest incomes, followed by Whites, while Black individuals and those from "other" racial groups earn the least. These differences can be attributed to historical and systemic factors, such as racial discrimination in the labor market and the underrepresentation of minorities in high-paying industries [@race]. Educational disparities between racial groups, as shown in @fig-Education-Race, also contribute to these income gaps.

Gender and race are exogenous factors beyond individual control. Addressing the gender pay gap requires targeted policies, such as promoting pay transparency and increasing women's participation in high-paying industries. To reduce racial income disparities, policymakers can implement measures like providing education and employment support for minority groups, encouraging diversity hiring, and strictly enforcing anti-discrimination laws. Cultural change is also crucial; reducing biases against women and racial groups can help narrow wage gaps and promote fairness.

@@

Income increases with work hours, but returns diminish. Beyond 50 hours per week, income becomes more unstable. This suggests that working longer hours does not always lead to higher pay. Policies like overtime pay rules or limits on work hours may play a role. The government should introduce better labor laws, such as capping excessive work hours. Encouraging businesses to innovate and improve efficiency can reduce reliance on long hours. These measures can protect workers’ rights while promoting a fairer and healthier labor market.

@@

This study's data has certain limitations. It relies on survey responses, which can include self-reporting bias. For example, respondents may overestimate or underestimate their income or working hours. Additionally, the original data collection only classified gender as male or female. Non-binary and transgender groups were not included, leaving some populations unrepresented. Future research could incorporate more diverse datasets, such as administrative records or third-party data, to improve accuracy. Using more detailed classifications could also better reflect overlooked groups and enhance inclusivity.

Some important variables were not included in the analysis, such as industry classification. While industry significantly affects income, its complex structure and lack of standardized categories made it difficult to analyze. This exclusion might have introduced bias into the results. To simplify the model, smaller racial groups, such as Indigenous, and multiracial individuals, were categorized as "Other." While this approach simplified the analysis, it overlooked finer racial differences. Future studies could refine these categories and include more detailed variables when sample sizes allow.
